
---
title: "Scribbles in Conjunction with the r-statistics.co by Selva Prabhakaran | Outlier Treatment"
output:
 html_document:
    df_print: paged
    code_folding: hide
    toc: TRUE
    toc_depth: 4
    toc_float: true
    theme: journal
    highlight: zenburn
    smart: false
editor_options: 
  chunk_output_type: inline
---

Hello World

Author: OA Lab, NWFSC
Title: Scribbles in Conjunction with the r-statistics.co by Selva Prabhakaran | Outlier Treatment Tutorial
Date: December 2020


# Version Check
```{r 0.1 Version Check , echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
#*********************************
## Version Check
#********************************* 
R.version

```


# Libraries
```{r 0.0 Libraries , echo=FALSE, results=FALSE, message=FALSE, warning=FALSE}
#*********************************
##Libraries
#********************************* 
library(stringr)
library(tidyverse)
library(plyr)
library(nlme)
library(tidyr)
library(dbplyr)
library(purrr)
library(wql)
library(lubridate)
library(tibbletime)
library(arsenal)
library(compareDF)
#for graphing
library(ggplot2)
library(ggfortify)
library(stringr)
library(nlme)
library(RColorBrewer)
#statistical analysis
library(gdata)
library(rsq)
library(doBy)
#Rnotebooks 
library(gridExtra)
library(kableExtra)

```



# Outlier Treatment
Outliers in data can distort predictions and affect the accuracy, 
if you don’t detect and handle them appropriately especially in regression models.

# Why outliers detection is important?
Treating or altering the outlier/extreme values in genuine observations 
is not a standard operating procedure. However, it is essential to 
understand their impact on your predictive models. It is left to the 
best judgement of the investigator to decide whether treating outliers 
is necessary and how to go about it.

So, why identifying the extreme values is important? Because, it can 
drastically bias/change the fit estimates and predictions. Let me 
illustrate this using the cars dataset.

To better understand the implications of outliers better, I am going to 
compare the fit of a simple linear regression model on cars dataset with 
and without outliers. In order to distinguish the effect clearly, 
I manually introduce extreme values to the original cars dataset. Then, 
I predict on both the datasets.

# 1. Inject outliers into data.
```{r # 1. Inject outliers into data.}
# Inject outliers into data.
cars1 <- cars[1:30, ]  # original data
cars_outliers <- data.frame(speed=c(19,19,20,20,20), dist=c(190, 186, 210, 220, 218))  # introduce outliers.
cars2 <- rbind(cars1, cars_outliers)  # data with outliers.
```



# 2. Plot of data with outliers
```{r 2. Plot of data with outliers}
# Plot of data with outliers.
par(mfrow=c(1, 2))
plot(cars2$speed, cars2$dist, xlim=c(0, 28), ylim=c(0, 230), main="With Outliers", xlab="speed", ylab="dist", pch="*", col="red", cex=2)
abline(lm(dist ~ speed, data=cars2), col="blue", lwd=3, lty=2)
```



# 3. Plot of original data without outliers
```{r 3. Plot of original data without outliers}
# Plot of original data without outliers. Note the change in slope (angle) of best fit line.
plot(cars1$speed, cars1$dist, xlim=c(0, 28), ylim=c(0, 230), main="Outliers removed \n A much better fit!", xlab="speed", ylab="dist", pch="*", col="red", cex=2)
abline(lm(dist ~ speed, data=cars1), col="blue", lwd=3, lty=2)
```
























```{r}
#**************E*N*D*************# 
#*********************************
## END OF SCRIPT | END OF DOCUMENT 
#*********************************
```


## END OF SCRIPT | END OF DOCUMENT



